---
title: "response_distribution"
author: "sunny yu"
date: "2026-04-17"
output: html_document
---

```{r setup, include=FALSE}
library(jsonlite)
library(dplyr)
library(purrr)
library(tidyr)
library(readr)
library(tibble)
library(stringr)
```



### DATA PREPROCESSING
## pre-process model response to show the image number for each option chosen

```{r cars}
## maps predicted label back to the image number and update the df


normalize_option <- function(x) {
  x %>%
    as.character() %>%
    str_remove_all("-") %>%
    str_to_lower()
}

# helper function for general cleaning
clean_item_uid <- function(df) {
  df %>%
    mutate(
      item_uid = as.character(item_uid),
      task_id = str_remove(item_uid, "_.*"),

      # vocab
      item_uid = str_replace(item_uid, "_word_", "__"),

      # theory of mind
      item_uid = if_else(
        task_id == "tom",
        str_remove(item_uid, "story[0-9]_"),
        item_uid
      ),
      item_uid = if_else(
        task_id == "tom",
        str_replace(item_uid, "-2", "_"),
        item_uid
      )
    ) %>%
    select(-task_id)
}

# helper for mental rotation lookup only
make_join_uid <- function(x, task_name) {
  if (task_name == "mental-rotation") {
    str_remove(x, "_[^_]+-[^_]+$")
  } else {
    x
  }
}

process_task_predictions <- function(task_name) {
  allowed_tasks <- c(
    "egma-math",
    "matrix-reasoning",
    "mental-rotation",
    "theory-of-mind",
    "trog",
    "vocab"
  )

  if (!task_name %in% allowed_tasks) {
    stop("task_name must be one of: ", paste(allowed_tasks, collapse = ", "))
  }

  task_prefix <- gsub("-", "_", task_name)
  simulated_df_name <- paste0(task_prefix, "_simulated")
  option_key_df_name <- paste0(task_prefix, "_option_key")

  if (!exists(simulated_df_name, envir = .GlobalEnv)) {
    stop("Object not found: ", simulated_df_name)
  }
  if (!exists(option_key_df_name, envir = .GlobalEnv)) {
    stop("Object not found: ", option_key_df_name)
  }

  simulated_df <- get(simulated_df_name, envir = .GlobalEnv)
  option_key_df <- get(option_key_df_name, envir = .GlobalEnv)

  option_orders <- fromJSON(
    "/Users/sunnyyu/Desktop/research/levante/data/option_orders_default.json",
    simplifyVector = FALSE
  )

  task_trials <- option_orders$tasks[[task_name]]$trials

  option_lookup <- map_dfr(task_trials, function(trial) {
    tibble(
      item_uid = trial$item_uid,
      label = map_chr(trial$option_order, "label"),
      option = map_chr(trial$option_order, "option")
    )
  })

  # special case: vocab item_uid mismatch between JSON and option key
  if (task_name == "vocab") {
    simulated_df <- simulated_df %>%
      mutate(item_uid = sub("^vocab__", "vocab_word_", item_uid))

    option_lookup <- option_lookup %>%
      mutate(item_uid = sub("^vocab__", "vocab_word_", item_uid))
  }

  simulated_df <- simulated_df %>%
    clean_item_uid() %>%
    mutate(
      predicted_label = as.character(predicted_label),
      join_uid = make_join_uid(item_uid, task_name)
    )

  option_lookup <- option_lookup %>%
    clean_item_uid() %>%
    mutate(
      label = as.character(label),
      option = as.character(option),
      join_uid = make_join_uid(item_uid, task_name)
    )

  option_key_df <- option_key_df %>%
    clean_item_uid() %>%
    mutate(
      across(starts_with("image"), as.character),
      join_uid = make_join_uid(item_uid, task_name)
    )

  processed_df <- simulated_df %>%
    left_join(
      option_lookup %>%
        select(join_uid, label, option),
      by = c("join_uid", "predicted_label" = "label")
    ) %>%
    rename(predicted_option = option) %>%
    left_join(
      option_key_df %>%
        select(join_uid, starts_with("image")),
      by = "join_uid"
    ) %>%
    select(-join_uid)

processed_df <- processed_df %>%
  mutate(
    predicted_option_match = normalize_option(predicted_option),
    image1_match = normalize_option(image1),
    image2_match = normalize_option(image2),
    image3_match = normalize_option(image3),
    image4_match = if ("image4" %in% names(.)) normalize_option(image4) else NA_character_
  )

if (task_name == "mental-rotation") {
  processed_df <- processed_df %>%
    mutate(
      matched_image = case_when(
        predicted_option_match == image1_match ~ "image1",
        predicted_option_match == image2_match ~ "image2",
        predicted_option_match == image3_match ~ "image3",
        TRUE ~ NA_character_
      )
    )
} else {
  processed_df <- processed_df %>%
    mutate(
      matched_image = case_when(
        predicted_option_match == image1_match ~ "image1",
        predicted_option_match == image2_match ~ "image2",
        predicted_option_match == image3_match ~ "image3",
        predicted_option_match == image4_match ~ "image4",
        TRUE ~ NA_character_
      )
    )
}


  processed_df %>%
    select(any_of(c(
      "item_uid", "predicted_label", "predicted_option",
      "image1", "image2", "image3", "image4", "matched_image"
    ))) %>%
    print(n = 20)

  output_file <- paste0(task_name, "_processed.csv")
  write_csv(processed_df, output_file)

  message("Saved processed file to: ", output_file)

  processed_df
}

```

```{r}
process_task_predictions("egma-math")
```






# process the processed_df above into a distribution df (same format as the human distribution)
```{r}

make_item_distribution <- function(processed_df, output_file = "distribution.csv") {
  
  # detect which image columns are present in matched_image
  possible_images <- c("image1", "image2", "image3", "image4")
  
  distribution_df <- processed_df %>%
    filter(!is.na(matched_image)) %>%
    
    # count occurrences
    count(item_uid, matched_image) %>%
    
    # reshape wide
    pivot_wider(
      names_from = matched_image,
      values_from = n,
      values_fill = 0
    )
  
  # add any missing image columns as 0
  for (img in possible_images) {
    if (!img %in% names(distribution_df)) {
      distribution_df[[img]] <- 0
    }
  }
  
  # normalize rows to sum to 1
  distribution_df <- distribution_df %>%
  mutate(across(all_of(possible_images), ~coalesce(.x, 0))) %>%
  rowwise() %>%
  mutate(
    total = sum(c_across(all_of(possible_images)))
  ) %>%
  ungroup() %>%
  mutate(
    across(all_of(possible_images), ~ ifelse(total > 0, .x / total, 0))
  ) %>%
  select(item_uid, all_of(possible_images))
  
  # save
  write_csv(distribution_df, output_file)
  message("Saved distribution file to: ", output_file)
  
  return(distribution_df)
}


```


```{r}
egma_math_distribution <- make_item_distribution(
  egma_math_processed,
  output_file = "egma-math_distribution.csv"
)


```

```{r}
matrix_reasoning_distribution <- make_item_distribution(
  matrix_reasoning_processed,
  output_file = "matrix-reasoning_distribution.csv"
)

```
```{r}
trog_distribution <- make_item_distribution(
  trog_processed,
  output_file = "trog_distribution.csv"
)
```


```{r}
vocab_distribution <- make_item_distribution(
  vocab_processed,
  output_file = "vocab_distribution.csv"
)
```


```{r}

tom_distribution <- make_item_distribution(
  theory_of_mind_processed,
  output_file = "theory-of-mind_distribution.csv"
)

```
```{r}
mental_rotation_distribution <- make_item_distribution(
  mental_rotation_processed,
  output_file = "mental_rotation_distribution.csv"
)

```


### distribution analysis 
## get KL divergence


```{r pressure, echo=FALSE}

# KL divergence for two probability vectors
kl_divergence <- function(p, q, epsilon = 1e-12) {
  p <- as.numeric(p)
  q <- as.numeric(q)
  
  # add epsilon for numerical stability
  p <- p + epsilon
  q <- q + epsilon
  
  # renormalize
  p <- p / sum(p)
  q <- q / sum(q)
  
  sum(p * log(p / q))
}

# main function
get_kl_by_item <- function(human_df, model_df, epsilon = 1e-12) {
  
  prob_cols <- c("image1", "image2", "image3", "image4")
  
  # keep only needed columns from model df
  model_small <- model_df %>%
    select(item_uid, all_of(prob_cols))
  
  # join model distribution onto every human row by item_uid
  joined_df <- human_df %>%
    left_join(
      model_small,
      by = "item_uid",
      suffix = c("_human", "_model")
    )
  
  # compute KL(human || model) for each human row
  kl_df <- joined_df %>%
    rowwise() %>%
    mutate(
      kl_divergence = kl_divergence(
        p = c_across(all_of(paste0(prob_cols, "_human"))),
        q = c_across(all_of(paste0(prob_cols, "_model"))),
        epsilon = epsilon
      )
    ) %>%
    ungroup()
  
  return(kl_df)
}
```



```{r}
kl_results <- get_kl_by_item(
  human_df = mental_rotation_proportions,
  model_df = mental_rotation_distribution
)
kl_results
write_csv(kl_results, 'mental_rotation_kl.csv')
```
## plotting code
# for each task, most similar to least similar task type 
```{r}
library(dplyr)
library(stringr)
library(ggplot2)

plot_df <- kl_results %>%
  mutate(
    item_type = str_extract(item_uid, "^[^_]+_[^_]+") # use this for egma-math, trog
    #item_type = str_extract(item_uid, "^[^_]+_[^_]+_[^_]+") # matrix-reasoning, vocab
  ) %>%
  group_by(item_type) %>%
  summarise(
    mean_kl = mean(kl_divergence, na.rm = TRUE),
    n = sum(!is.na(kl_divergence)),
    .groups = "drop"
  ) %>%
  arrange(mean_kl) %>%  # lowest KL = most similar
  mutate(
    item_type = factor(item_type, levels = item_type)
  )

ggplot(plot_df, aes(x = item_type, y = mean_kl)) +
  geom_col() +
  coord_flip() +
  labs(
    x = "Item type",
    y = "Average KL divergence",
    title = "Average KL divergence by item type (vocab)",
    subtitle = "Ranked from most similar to least similar"
  ) +
  theme_classic(base_size = 14)
```



# average kl_divergence for each ability bin (averaged across task)
```{r}

# average KL per ability bin (across all items/tasks)
plot_df <- kl_results %>%
  group_by(ability_bin) %>%
  summarise(
    mean_kl = mean(kl_divergence, na.rm = TRUE),
    n = sum(!is.na(kl_divergence)),
    .groups = "drop"
  ) %>%
  arrange(mean_kl) %>%  # most similar (low KL) → least similar
  mutate(
    ability_bin = factor(ability_bin, levels = ability_bin)
  )

plot_df <- plot_df %>%
  mutate(
    bin_start = as.numeric(sub("_.*", "", ability_bin))
  ) %>%
  arrange(bin_start) %>%
  mutate(ability_bin = factor(ability_bin, levels = ability_bin))

# plot
ggplot(plot_df, aes(x = ability_bin, y = mean_kl)) +
  geom_col() +
  labs(
    x = "Ability bin",
    y = "Average KL divergence",
    title = "Average KL divergence by ability bin (vocab, overall)",
    subtitle = "Averaged across all tasks (ranked)"
  ) +
  theme_classic(base_size = 14)
```
# average KL per bin per task

```{r}

# extract task type + compute mean KL per (task_type, ability_bin)
plot_df <- kl_results %>%
  mutate(
    task_type = str_extract(item_uid, "^[^_]+_[^_]+")  #math_add
  ) %>%
  filter(task_type %in% c(
   # "math_multiply", "math_identify", "math_add",
  #  "math_missing", "math_subtract", "math_compare"
    "trog_2combo", "trog_3combo"
  )) %>%
  group_by(task_type, ability_bin) %>%
  summarise(
    mean_kl = mean(kl_divergence, na.rm = TRUE),
    n = sum(!is.na(kl_divergence)),
    .groups = "drop"
  )

# optional: order ability bins numerically (e.g., -1_0, 0_1, ...)
plot_df <- plot_df %>%
  mutate(
    bin_start = as.numeric(sub("_.*", "", ability_bin))
  ) %>%
  arrange(task_type, bin_start) %>%
  group_by(task_type) %>%
  mutate(
    ability_bin = factor(ability_bin, levels = unique(ability_bin))
  ) %>%
  ungroup()

# facet plot (one panel per task type)
ggplot(plot_df, aes(x = ability_bin, y = mean_kl)) +
  geom_col() +
  facet_wrap(~ task_type, scales = "free_y") +
  labs(
    x = "Ability bin",
    y = "Average KL divergence",
    title = "Average KL divergence by ability bin and task type (trog)"
  ) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}
# extract task type + compute mean KL per (task_type, ability_bin)
plot_df <- kl_results %>%
  mutate(
    task_type = str_extract(item_uid, "^[^_]+_[^_]+_[^_]+")  #math_add
  ) %>%
  filter(task_type %in% c(
    "matrix_set3_md", "matrix_set3_pd"
  )) %>%
  group_by(task_type, ability_bin) %>%
  summarise(
    mean_kl = mean(kl_divergence, na.rm = TRUE),
    n = sum(!is.na(kl_divergence)),
    .groups = "drop"
  )

# optional: order ability bins numerically (e.g., -1_0, 0_1, ...)
plot_df <- plot_df %>%
  mutate(
    bin_start = as.numeric(sub("_.*", "", ability_bin))
  ) %>%
  arrange(task_type, bin_start) %>%
  group_by(task_type) %>%
  mutate(
    ability_bin = factor(ability_bin, levels = unique(ability_bin))
  ) %>%
  ungroup()

# facet plot (one panel per task type)
ggplot(plot_df, aes(x = ability_bin, y = mean_kl)) +
  geom_col() +
  facet_wrap(~ task_type, scales = "free_y") +
  labs(
    x = "Ability bin",
    y = "Average KL divergence",
    title = "Average KL divergence by ability bin and task type (matrix reasoning)"
  ) +
  theme_classic(base_size = 14) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```



